Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-01 manuscript.tex
Update website url
hyperlinks, section on reproducing, split long eqs

\documentclass[paper,twocolumn]{geophysics}
%\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\title{
Tesseroids: forward modeling gravitational fields in spherical coordinates
}
\righthead{Tesseroids}
\ms{Submission} % manuscript number

\author{
    Leonardo Uieda\footnotemark[1]\footnotemark[2],
    et al.
}

\address{
    \footnotemark[1]Universidade do Estado do Rio de Janeiro,
    Rio de Janeiro, Brazil.
    email: leouieda@gmail.com
    \footnotemark[2]Observat\'orio Nacional,
    Rio de Janeiro, Brazil.
}

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$ inside the tesseroid,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

@@

\begin{equation}
    V(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{1}{\ell}
        \kappa\  dr^\prime d\phi^\prime d\lambda^\prime,
    \label{eq:tesspot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{\Delta_\alpha}{\ell^3}
        \kappa\ dr^\prime d\phi^\prime d\lambda^\prime,
    \label{eq:tessgrav}
\end{equation}
\noindent
and
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \left(
            \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
            \frac{\delta_{\alpha\beta}}{\ell^3}
        \right)
        \kappa\ dr^\prime d\phi^\prime d\lambda^\prime.
    \label{eq:tesstensor}
\end{equation}

@@

\begin{eqnarray}
    \Delta_x &=& r^\prime\cos\phi\sin\phi^\prime - \sin\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda), \\
    \Delta_y &=& r^\prime \cos \phi^\prime \sin(\lambda^\prime - \lambda), \\
    \Delta_z &=& r^\prime \cos \psi - r,\\
    \kappa &=& {r^\prime}^2 \cos \phi^\prime, \\
    \ell &=& \sqrt{{r^\prime}^2 + r^2 - 2 r^\prime r \cos \psi}, \\
    \cos\psi &=& \sin\phi\sin\phi^\prime + \cos\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda).
\end{eqnarray}

@@

\begin{equation}
    \iiint\limits_{\Omega}
    f(r^\prime, \lambda^\prime, \phi^\prime)
    dr^\prime d\lambda^\prime d\phi^\prime
    \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    f(r_i, \phi_j, \lambda_k),
    \label{eq:glq3d}
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point mases instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30,
        the result is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

We have implemented the calculation of
the tesseroid gravitational fields
with adaptive discretization
in the open-source package \textit{Tesseroids}.
The software is freely available
online\footnote{http://www.leouieda.com/tesseroids}
under the BSD 3-clause open-source license.
An archived version of the source code
is available as part of this article.

@@

All figures and numerical experiments in this article
were produced by IPython notebooks\footnote{http://ipython.org}
\citep{Perez2007}.
These files combine the analysis and figure generation source code
with text, equations, and the figures generated by the code.
These files,
as well as instructions for installing and running the analysis,
are distributed along with the source code archive
that accompanies this article.
Alternatively,
the notebooks and can also be downloaded
or viewed in a non-executable form from
an online repository\footnote{https://github.com/pinga-lab/paper-tesseroids}.

@@

\begin{equation}
    d = \left[
        r^2 + r_t^2 - 2 r r_t
        \cos(\sin\phi\sin\phi_t + \cos\phi\cos\phi_t
            \cos(\lambda - \lambda_t))
        \right]^{\frac{1}{2}}.
    \label{eq:distance}
\end{equation}

@@

@@

Figure~\ref{fig:division} shows examples of
the resulting tesseroid models after adaptive discretization.
Figure~\ref{fig:division}a shows
the initial tesseroid and computation point P.
Figures~\ref{fig:division}b-d are
the result of adaptive discretization using
different values of the distance-size ratio $D$,
respectively,
$D=1$, $D=2$, and $D=6$.
The number of tesseroids in the resulting discretization is, respectively,
4, 38, and 936.
Code to perform the adaptive discretization
and produce Figure~\ref{fig:division} is available in
the IPython notebook \texttt{methods\_figures.ipynb}.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

>>>>

\pdfoutput=1
%\documentclass[paper,twocolumn]{geophysics}
%\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
\documentclass[twocolumn]{article}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

\hypersetup{
    colorlinks=true,
    citecolor=blue,
    linkcolor=blue,
    urlcolor=blue,
    pdfauthor={Uieda et al.},
    pdftitle={Tesseroids: forward modeling gravitational fields in spherical
              coordinates}
}
\title{
    Tesseroids: forward modeling gravitational fields in spherical coordinates
}
\author{
    Leonardo Uieda\footnotemark[1]\footnotemark[2],
    et al.
}

@@

@@

%\righthead{Tesseroids}
%\ms{Submission} % manuscript number

%\address{
    \noindent
    \footnotemark[1]Universidade do Estado do Rio de Janeiro,
    Rio de Janeiro, Brazil.
    email: leouieda@gmail.com
    \footnotemark[2]Observat\'orio Nacional,
    Rio de Janeiro, Brazil.
%}

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$ inside the tesseroid,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

@@

\begin{equation}
    V(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{1}{\ell}
        \kappa\  dr^\prime d\phi^\prime d\lambda^\prime,
    \label{eq:tesspot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{\Delta_\alpha}{\ell^3}
        \kappa\ dr^\prime d\phi^\prime d\lambda^\prime,
    \label{eq:tessgrav}
\end{equation}
\noindent
and
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        I_{\alpha\beta}
        \ \kappa\ dr^\prime d\phi^\prime d\lambda^\prime.
    \label{eq:tesstensor}
\end{equation}
\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right)
    \label{eq:tesstensorkernel}
\end{equation}

@@

\begin{equation}
    \Delta_x = r^\prime(\cos\phi\sin\phi^\prime - \sin\phi\cos\phi^\prime
               \cos(\lambda^\prime - \lambda)),
\end{equation}
\begin{equation}
    \Delta_y = r^\prime \cos \phi^\prime \sin(\lambda^\prime - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r^\prime \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r^\prime}^2 \cos \phi^\prime,
\end{equation}
\begin{equation}
    \ell = \sqrt{{r^\prime}^2 + r^2 - 2 r^\prime r \cos \psi},
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi^\prime + \cos\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda).
\end{equation}

@@

\begin{equation}
    \iiint\limits_{\Omega}
    f(r^\prime, \lambda^\prime, \phi^\prime)
    d\Omega
    \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    f(r_i, \phi_j, \lambda_k),
    \label{eq:glq3d}
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\columnwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point mases instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30,
        the result is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

We have implemented the calculation of
the tesseroid gravitational fields
with adaptive discretization
in the open-source package \textit{Tesseroids}.
The software is freely available
online\footnote{
\href{http://tesseroids.leouieda.com}{http://tesseroids.leouieda.com}}
under the BSD 3-clause open-source license.
An archived version of the source code
is available as part of this article.

@@

\subsection{Reproducing analysis and results}

All figures and numerical experiments in this article
were produced by IPython notebooks\footnote{
\href{http://ipython.org}{http://ipython.org}}
\citep{Perez2007}.
These files combine the analysis and figure generation source code
with text, equations, and the figures generated by the code.
The notebook files,
as well as instructions for installing and running the analysis,
are distributed along with the source code archive
that accompanies this article.
Alternatively,
all accompanying material can be downloaded
from an online source code repository\footnote{
\href{https://github.com/pinga-lab/paper-tesseroids}{https://github.com/pinga-lab/paper-tesseroids}}
or a persistent archive\footnote{
\href{}{Zenodo DOI}}.

@@

\begin{equation}
    d = \left[
        r^2 + r_t^2 - 2 r r_t \cos\psi_t
        \right]^{\frac{1}{2}}.
    \label{eq:distance}
\end{equation}
\begin{equation}
    \cos\psi_t =
        \sin\phi\sin\phi_t + \cos\phi\cos\phi_t\cos(\lambda - \lambda_t)
\end{equation}

@@

\noindent
$L_\lambda$ and $L_\phi$ are arc-distances measured along the top surface of
the tesseroid (Figure~\ref{fig:division}a).
Specifically, $L_\lambda$ is measured long the middle latitude of the
tesseroid ($\phi_t$).

@@

Figure~\ref{fig:division} shows examples of
the resulting tesseroid models after adaptive discretization.
Figure~\ref{fig:division}a shows
the initial tesseroid and computation point P.
Figures~\ref{fig:division}b-d are
the result of adaptive discretization using
different values of the distance-size ratio $D$,
respectively,
$D=1$, $D=2$, and $D=6$.
The number of tesseroids in the resulting discretization is, respectively,
4, 38, and 936.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}
